A thermodynamic view of dusty protoplanetary disks

### 2017-07-02 linear.tex

\section{Linearized equations}\label{linear_problem}
To compute explicit solutions to the linear problem, 
we consider radially-localized axisymmetric disturbances of the form  
\begin{align}
  \delta X (r, z) = \delta X_1(r,z)\exp{(\ii k_x r)},
\end{align} 
%and similarly for $\dd P$ and $\dd\bm{v}$. 
where $k_x$ is a real wavenumber such that $|k_xr|\gg 1$, and the 
amplitude $\dd X_1(r,z)$ is 
a slowly-varying function of $r$. Then 
$\p_r\to i k_x$ when acting on the above primitive perturbations, and we may
neglect curvature terms. We take  
$k_x>0$ without loss of generality. Hereafter, we drop the subscript 1
on the amplitudes. 
%The frequency $\sigma = \omega +
%\ii s$ is generally complex, with $\omega$ being the real frequency
%and $s$ is the real growth rate.

@@

\begin{align}
  \ii\sigma W &= \ii k_x \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi 
% +  \delta\bm{F}\cdot\hat{\bm{r}}
- W F_r - \ii k_x Q - \ii k_x\dd\psi,\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &= - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right] - \dd\psi^\prime  %\delta\bm{F}\cdot\hat{\bm{z}} 
,\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k_x \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2} %, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy}\\
\dd\psi^{\prime\prime}  &= 4\pi G \rho W + k_x^2\dd\psi, \label{lin_sg}
\end{align}  
where $^\prime \equiv \p_z$ and recall $\bm{F} \equiv -\nabla
P/\rho$. Here we restore self-gravity temporarily for SGI in the next
section. The linearized dust-duffusion function 
$\dd\mathcal{C}$ is given in  Appendix \ref{lin_dust}. 
% and $\dd\bm{F}$ is the linearized pressure
%force, given in Appendix \ref{lin_press}. 
%Note that we have assumed a temperature profile that only depends on $r$.

@@

\subsection{Secular gravitational instability}\label{sgi}
We first consider a razor-thin, self-gravitating disk so that $\rho =
\Sigma\delta(z)$. Here $\delta$ is the Dirac delta function and $\Sigma$ is
the total surface density. Now $\epsilon$ and 
$\tepsilon$ refers to the global dust-to-gas ratio and dust 
fraction, respectively. The background disk is
uniform and we neglect the vertical dimension   
in Eq. \ref{lin_mass}---\ref{lin_energy}. The linearized dust-gas drag
term is then $-\delta\mathcal{C}/\rho = \tstop\tepsilon c_s^2k_x^2
Q$. The thin-disk solution to Eq. \ref{lin_sg} is $\dd\psi(z=0) = -2\pi G
\Sigma W/\left|k_x\right|$.%, where $\Sigma$ is the total surface
%density.

@@

\subsection{Streaming instability}\label{si}
We now consider 3D disks without self-gravity. We neglect the vertical     
component of the stellar gravity, appropriate for studying regions
near the disk midplane. This  
allows us to Fourier analyze in $z$ to obtain 
an algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability (SI) in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

>>>>

\section{Application to protoplanetary disks}\label{linear_problem}
We now examine instabilities in dusty protoplanetary disks
based on explicit descriptions of the equilibrium state. We first show
that sharp radial edges in the dust component are unstable. We then
revisit two well-known instabilities in dusty gas using the
hydrodynamic framework developed thusfar, namly secular graviational
instabilities and the streaming instability.

\subsection{Disk structure with a prescribed dust distribution}\label{steady_state}  
We assume a Gaussian profile in the
dust-to-gas ratio,    
\begin{align}\label{dust_gauss}
  \epsilon(r,z) = \epsilon_0(r)
  \exp{\left[-\frac{z^2}{2\Htilde^2(r)}\right]}. 
\end{align}
%For simplicity we assume the dust-to-gas ratio at the mid-plane
%$\epsilon_0$, as well as its characteristic scale-height
%$\widetilde{H}$, are both constant.
% We take the mid-plane dust-to-gas ratio
%to be a power-law in radius,  
%\begin{align}
%  \epsilon_0(r) = \epsilon_{00}\left(\frac{r}{r_0}\right)^{-d},  
%\end{align}
%where $\epsilon_{00}$ is the dust-to-gas ratio at the fiducial radius
%$r_0$.  
Inserting Eq. \ref{dust_gauss} into vertical hydrostatic equilibrium,
Eq. \ref{steady_momz} and integrating with the approximate
gravitational potential (Eq. \ref{thin_disk_potential}) we obtain the
gas density as
\begin{align}
  &\rhog(r,z)= \notag\\
&\rho_\mathrm{g0}(r)\exp{\left\{ - \frac{z^2}{2\Hgas^2}
    -\epsilon_0\frac{\Htilde^2}{\Hgas^2}\left[1 -
      \exp{\left(-\frac{z^2}{2\Htilde^2}\right)}\right] \right\}}, 
\end{align}
where
\begin{align}
  \Hgas = \frac{c_s}{\OmK}, \quad \OmK \equiv \sqrt{\frac{GM_*}{r^3}},   
\end{align}
is the gas scale-height in the dust-free limit and $\OmK$ is the
Keplerian frequency, respectively.

We usually consider gas-dominated disks with $\epsilon_0 < 1$.  
Then $\rhog(r,z)$ is effectively Gaussian, as in the 
dust-free case. The dust density is approximately 
\begin{align}
  \rhod \simeq \epsilon_0\rho_\mathrm{g0}(r) \exp
        {\left(-\frac{z^2}{2H_\mathrm{d}^2}\right)}, 
\end{align}
with 
\begin{align}
  \frac{1}{H_\mathrm{d}^2} = \frac{1}{\Htilde^2} + \frac{1}{\Hgas^2}, 
\end{align}
and $H_\mathrm{d}$ is the dust-scale height. In numerical
calculations, we  we specify $H_\mathrm{d}< \Hgas$ to obtain 
$\Htilde$ for input.

Finally, we define 
\begin{align}
  Z \equiv \epsilon_0\frac{\Hd}{\Hg} \simeq
  \frac{\Sigma_\mathrm{d}}{\Sigma_\mathrm{g}} 
\end{align}
as a measure of the local metalicity, where $\Sigma_\mathrm{d}$ and
$\Sigma_\mathrm{g}$ are the dust and gas surface densities,
respectively. The second equality holds for $\epsilon_0\ll1$.

%\Sigma_\mathrm{d}$ and the gas
%surface density $\Sigma_\mathrm{g}$.

\subsubsection{Orbital frequency} 
In the thin-disk approximation, the disk orbital frequency is 
\begin{align}
  \Omega(r,z) = \OmK(r)\left[1 - \frac{3}{2}\frac{z^2}{r^2} +
    \frac{h_\mathrm{g}^2}{\left(1+\epsilon\right)}\frac{\p}{\p\ln{r}}\ln{\left(c_s^2\rhog\right)}
    \right]^{1/2}, 
\end{align}
where 
\begin{align}
  h_\mathrm{g} \equiv \frac{\Hgas}{r}
\end{align}
is the characteristic disk aspect-ratio.

\subsubsection{Vertical shear}\label{vertshear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p \seff}{\p z} -\frac{\p
    P}{\p z}\frac{\p \seff}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align}\label{vshear2}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
%  -\frac{\p\epsilon}{\p z}\left[\frac{\p\ln{\rhog}}{\p r} + \frac{q}{r}\right]\right.\notag\\
 -\frac{\p\epsilon}{\p z}\frac{\p\ln{P}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r} \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}
  \right\}. 
\end{align}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio. The third term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient, and survives in the dust-free limit.

We can compare these sources by 
evaluating them using the equilibrium
solutions in \S\ref{steady_state}. We assume the disk is radially
smooth so that $\p_r\sim 1/r$. % and the dust-to-gas ratio
% $\epsilon\ll1$. 
This gives

\begin{align}\label{vshear_split}
  \frac{\left|r\p_z\Omega\right|_{\text{
        dust/gas gradient}}}{\left|r\p_z\Omega\right|_{\text{
        temp. gradient}}} \sim
 % \epsilon \frac{\mathrm{max}\left(\delta^2,
  %  1\right)}{\left|q\right|\left(\epsilon + \delta^2\right)},
 \epsilon \frac{\mathrm{max}\left(\delta^2,
    1\right)}{\left|q\right|\left(1+\epsilon\right)\delta^2},
\end{align}
where $\delta\equiv \Htilde/\Hgas$. 
Since $|q|=O(1)$ in PPDs, Eq. \ref{vshear_split} indicates that
vertical shear due to variations in the dust-to-gas ratio dominates 
over that due to the radial temperature gradient for thin dust layers
such that $\delta^2\ll \epsilon$. Otherwise, vertical shear is
associated with the radial temperature gradient. %However, axisymmtric
%\emph{instability} always caused by $\p_rc_s^2$.

\subsubsection{Dusty vertical buoyancy}\label{vbuoyancy}
For the above equilibria the vertical buoyancy frequency is given
explicitly as 
\begin{align}
  N_z^2 =
  %\frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
%  z}\frac{\p\epsilon}{\p z} \\ &
  %                                =
 \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
where we have used $c_s=c_s(r)$. Then 
\begin{align*}
N_z\lesssim
O\left(\sqrt{\epsilon_0}\OmK\right). 
\end{align*}
However, for well-mixed dust layers such that $H_\epsilon \gg \Hg$, 
$\mathrm{max}\left(N_z\right)$ may occur outside a finite vertical domain.

\subsection{Instability of dusty edges} 
Here we apply the dusty analog of the Solberg-Hoiland criteria derived
in \S\ref{iso_perfect}. As discussed there, the first criterion in
generally satisifed. Thus we only consider the second condition,
Eq. \ref{dusty_solberg2}.

We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The second term is usually stabilizing in 
PPDs for reasons given in \S\ref{iso_perfect}. The third term is
always destabilizing, but is small in smooth, thin disks with radial
gradients $O(1/r)$, $h_\mathrm{g}\ll 1$, 
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occured). This means that typical dusty PPDs are stable to
axisymmetric perturbations, a conclusion valid for arbitrarily thin
dust layers.

%%%%%%%%%%%%%%%%%%

To violate Eq. \ref{stability_est2} and obtain instability, notice the 
LHS is a quadratic in $\p_r\epsilon$. Thus instability is possible for
sufficiently large (in magnitude) radial gradients  in the dust-to-gas
ratio,  
\begin{align}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} > \chi_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\epsilon}}{\p\ln{r}} < \chi_-,
\end{align}
for \emph{instability}, where
\begin{align}\label{spm}
\chi_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
  \left[
  \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
  4 \frac{\Htilde^2}{H_\mathrm{g}^2}
  \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align} 
In typical accretion disks where $\p_r\rhog<0$, instability is easier
for increasing dust-to-gas ratios ($\p_r\epsilon > 0$).

We can neglect density gradients in Eq. \ref{spm} 
if $r\p_r\ln{\rhog}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
5\times10^{-3}$. (So the dust layer thickness $\Hd\gg 5
\times10^{-3}\Hgas$ as well.) Then 
instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio 
is much less its vertical lengthscale, $L_\epsilon\ll O(\Htilde)$, 
then the system is  unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets.

Technically, the above discussion is only applicable to strictly
isothermal disks perfectly coupled to dust (see \S\ref{iso_perfect}). 
However, since dusty edges translate to sharp entropy gradients (see
\S\ref{dusty_entropy}), we may generally expect sharp features in the
dust distribution of protoplanetary disks to be unstable.

%{\bf mention this is equivalent to sharp entropy gradients?}

\subsection{Radially localized linear problem}
We now specialize further and compute explicit solutions to the linear
problem. We consider radially-localized axisymmetric disturbances of the form  
\begin{align}
  \delta X (r, z) = \delta X_1(r,z)\exp{(\ii k_x r)},
\end{align} 
%and similarly for $\dd P$ and $\dd\bm{v}$. 
where $k_x$ is a real wavenumber such that $|k_xr|\gg 1$, and the 
amplitude $\dd X_1(r,z)$ is 
a slowly-varying function of $r$. Then 
$\p_r\to i k_x$ when acting on the above primitive perturbations, and we may
neglect curvature terms. We take  
$k_x>0$ without loss of generality. Hereafter, we drop the subscript 1
on the amplitudes. 
%The frequency $\sigma = \omega +
%\ii s$ is generally complex, with $\omega$ being the real frequency
%and $s$ is the real growth rate.

@@

\begin{align}
  \ii\sigma W &= \ii k_x \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi 
% +  \delta\bm{F}\cdot\hat{\bm{r}}
- W F_r - \ii k_x Q - \ii k_x\dd\psi,\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &= - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right] - \dd\psi^\prime  %\delta\bm{F}\cdot\hat{\bm{z}} 
,\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k_x \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2} %, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy}\\
\dd\psi^{\prime\prime}  &= 4\pi G \rho W + k_x^2\dd\psi, \label{lin_sg}
\end{align}  
where $^\prime \equiv \p_z$ and recall $\bm{F} \equiv -\nabla
P/\rho$. We have temporarily restored self-gravity  to
discuss SGI in the next
section. The linearized dust-duffusion function 
$\dd\mathcal{C}$ is given in  Appendix \ref{lin_dust}. 
% and $\dd\bm{F}$ is the linearized pressure
%force, given in Appendix \ref{lin_press}. 
%Note that we have assumed a temperature profile that only depends on $r$.

@@

\subsubsection{Secular gravitational instability}\label{sgi}
We first consider a razor-thin, self-gravitating disk so that $\rho =
\Sigma\delta(z)$. Here $\delta$ is the Dirac delta function and $\Sigma$ is
the total surface density. Now $\epsilon$ and 
$\tepsilon$ refers to the global dust-to-gas ratio and dust 
fraction, respectively. The background disk is
uniform and we neglect the vertical dimension   
in Eq. \ref{lin_mass}---\ref{lin_energy}. The linearized dust-gas drag
term is then $-\delta\mathcal{C}/\rho = \tstop\tepsilon c_s^2k_x^2
Q$. The thin-disk solution to Eq. \ref{lin_sg} is $\dd\psi(z=0) = -2\pi G
\Sigma W/\left|k_x\right|$.%, where $\Sigma$ is the total surface
%density.

@@

\subsubsection{Streaming instability}\label{si}
We now consider 3D disks without self-gravity. We neglect the vertical     
component of the stellar gravity, appropriate for studying regions
near the disk midplane. This  
allows us to Fourier analyze in $z$ to obtain 
an algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability (SI) in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.
